---
title: "Untitled2"
output: html_document
date: "2025-02-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(immunarch)
library(ragg)
```

```{r}
#load(CRC_immdata.RData)
```

## R Markdown

# Load data and downsample

```{r}
#### PASTE immdata code here ####
```

# Add TNM Labels 

```{r}
labels <- read.csv("/home/ubuntu/CRC/CRC_Project/CRC_blood/CRC_blood_data/t_n_m_labels.csv", 
                   header = FALSE, 
                   stringsAsFactors = FALSE,
                   col.names = c("Sample", "Sample2", "T", "N", "M"))

# Step 2: Optionally drop the redundant column if it’s not needed.
labels <- labels[, c("Sample", "T", "N", "M")]

# (Optional) View the labels data frame to verify it looks correct.
head(labels)
```
```{r}
# Filter immdata$meta so it only has rows for the samples still in immdata$data
immdata$meta <- immdata$meta[immdata$meta$SampleID %in% names(immdata$data), ]

# Merge labels
immdata$meta <- merge(immdata$meta, labels, by.x = "SampleID", by.y = "Sample", all.x = TRUE)
```


# Add Stage
```{r}
# Copy the metadata
stage_metadata <- immdata$meta

stage_metadata <- stage_metadata %>%
  mutate(
    T_num = suppressWarnings(as.numeric(T)),
    N_num = suppressWarnings(as.numeric(N)),
    M_num = suppressWarnings(as.numeric(M))
  )
```

```{r}
stage_metadata <- stage_metadata %>%
  mutate(
    Stage = case_when(
      # Stage IV: any T, any N, M > 0
      !is.na(M_num) & M_num > 0 ~ "Stage IV",
      
      # Stage III: any T, N=1 or 2, M=0
      !is.na(N_num) & (N_num %in% c(1, 2)) & (!is.na(M_num) & M_num == 0) ~ "Stage III",
      
      # Stage II: T=3 or 4, N=0, M=0
      (T_num %in% c(3, 4)) & (!is.na(N_num) & N_num == 0) & (!is.na(M_num) & M_num == 0) ~ "Stage II",
      
      # Stage I: T=1 or 2, N=0, M=0
      (T_num %in% c(1, 2)) & (!is.na(N_num) & N_num == 0) & (!is.na(M_num) & M_num == 0) ~ "Stage I",
      
      # Everything else or incomplete: Unknown
      TRUE ~ "Unknown"
    )
  )
```


# Downsample 
```{r}
downsampledData = repSample(immdata$data, .method = "downsample")
```


############################################################
### Analyses with Stage (wihtout Unknown) ###
############################################################

1. unique
2. clonality (top, rare)
3. diversity (true diversity, inverse simpson, gini simpson)

```{r}
# Repertoire Volume (Number of Unique Clones)

# Filter out any samples where Stage is NA or labeled "Unknown"
stage_metadata_filtered <- stage_metadata[!is.na(stage_metadata$Stage), ]
stage_metadata_filtered <- subset(stage_metadata_filtered, Stage != "Unknown")
stage_metadata_filtered <- stage_metadata_filtered %>%
  rename(Sample = SampleID)

# Also filter the actual data so the sample names align
downsampledData_filtered <- downsampledData[names(downsampledData) %in% stage_metadata_filtered$Sample]
```


```{r}
# Now run your code with the filtered data
uniq <- repExplore(downsampledData_filtered, .method = "volume")
agg_png("rep_volume_stages_plot.png", width = 1200, height = 900, res = 144)
p_uniq <- vis(uniq, .by = "Stage", .meta = stage_metadata_filtered, .test = FALSE)
print(p_uniq)
dev.off()
```


```{r}
# Clonality (Top clones)
top <- repClonality(downsampledData_filtered, .method = "top", .head = c(10,100,1000))

ragg::agg_png("clonality_top_clones_stages_plot.png", width = 1200, height = 900, res = 144)
p_top <- vis(top, .by = "Stage", .meta = stage_metadata_filtered, .test = FALSE)
print(p_top)
dev.off()
```

```{r}
# Diversity (True diversity)
trued <- repDiversity(downsampledData_filtered, .method = "div", .verbose = FALSE)
ragg::agg_png("true_diversity_stages_plot.png", width = 1200, height = 900, res = 144)
p_trued <- vis(trued, .by = "Stage", .meta = stage_metadata_filtered, .test = FALSE)
print(p_trued)
dev.off()
```

# Other diversity indices:

```{r}
# "inv.simp" (Inverse Simpson), "gini.simp", etc.
inv <- repDiversity(downsampledData_filtered, .method = "inv.simp", .verbose = FALSE)
ragg::agg_png("inverse_simpson_diversity_stages_plot.png", width = 1200, height = 900, res = 144)
p_inv <- vis(inv, .by = "Stage", .meta = stage_metadata_filtered, .test = FALSE)
print(p_inv)
dev.off()


simp <- repDiversity(downsampledData_filtered, .method = "gini.simp", .verbose = FALSE)
ragg::agg_png("gini_simpson_diversity_stages_plot.png", width = 1200, height = 900, res = 144)
p_gini <- vis(simp, .by = "Stage", .meta = stage_metadata_filtered, .test = FALSE)
print(p_gini)
dev.off()
```

############################################################
### Analyses with Stage with Unknown ###
############################################################

1. unique
2. clonality (top, rare)
3. diversity (true diversity, inverse simpson, gini simpson)

```{r}
############################################################
### Analyses with Stage ###
############################################################
# Repertoire Volume (Number of Unique Clones)

# Filter out any samples where Stage is NA or labeled "Unknown"
stage_metadata_filt_unknown <- stage_metadata[!is.na(stage_metadata$Stage), ]
stage_metadata_filt_unknown <- stage_metadata_filt_unknown %>%
  rename(Sample = SampleID)

# Also filter the actual data so the sample names align
downsampledData_filt_unknown <- downsampledData[names(downsampledData) %in% stage_metadata_filt_unknown$Sample]
```


```{r}
# Now run your code with the filtered data
uniq <- repExplore(downsampledData_filt_unknown, .method = "volume")
agg_png("rep_volume_unkown_stages_plot.png", width = 1200, height = 900, res = 144)
p_uniq <- vis(uniq, .by = "Stage", .meta = stage_metadata_filt_unknown, .test = FALSE)
print(p_uniq)
dev.off()
```


```{r}
# Clonality (Top clones)
top <- repClonality(downsampledData_filt_unknown, .method = "top", .head = c(10,100,1000))

ragg::agg_png("clonality_top_clones_unknown_stages_plot.png", width = 1200, height = 900, res = 144)
p_top <- vis(top, .by = "Stage", .meta = stage_metadata_filt_unknown, .test = FALSE)
print(p_top)
dev.off()
```

```{r}
# Diversity (True diversity)
trued <- repDiversity(downsampledData_filt_unknown, .method = "div", .verbose = FALSE)
ragg::agg_png("true_diversity_unkown_stages_plot.png", width = 1200, height = 900, res = 144)
p_trued <- vis(trued, .by = "Stage", .meta = stage_metadata_filt_unknown, .test = FALSE)
print(p_trued)
dev.off()
```

# Other diversity indices:

```{r}
# "inv.simp" (Inverse Simpson), "gini.simp", etc.
inv <- repDiversity(downsampledData_filt_unknown, .method = "inv.simp", .verbose = FALSE)
ragg::agg_png("inverse_simpson_diversity_unknown_stages_plot.png", width = 1200, height = 900, res = 144)
p_inv <- vis(inv, .by = "Stage", .meta = stage_metadata_filt_unknown, .test = FALSE)
print(p_inv)
dev.off()


simp <- repDiversity(downsampledData_filt_unknown, .method = "gini.simp", .verbose = FALSE)
ragg::agg_png("gini_simpson_diversity_unknown_stages_plot.png", width = 1200, height = 900, res = 144)
p_gini <- vis(simp, .by = "Stage", .meta = stage_metadata_filt_unknown, .test = FALSE)
print(p_gini)
dev.off()
```

Present the Stages counts

```{r}
# dataframe with stage counts
stage_counts <- stage_metadata %>%
  group_by(Stage) %>%
  summarise(Count = n())

agg_png("stage_counts_with_labels.png", width = 1200, height = 900, res = 144)

# Create the ggplot 
p <- ggplot(stage_counts, aes(x = Stage, y = Count, fill = Stage)) +
  geom_bar(stat = "identity") +
  # Add text labels: label = Count and position above the bars
  geom_text(aes(label = Count), 
            vjust = -0.3,
            size = 4,             
            fontface = "plain") +  
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Count of Samples by Stage",
       x = "Stage",
       y = "Number of Samples") +
  theme_minimal()

print(p)

dev.off()
```


